Notes and other text

### 2023-04-19 doc/manuscript/translocation.qmd
Shorten Intro and Results and add frog count summary

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Allie comment. I believe evolutionary rescue also encompasses the introduction of new variation through gene flow. See box 1 in ref 16 for a helpful discussion of genetic vs evolutionary rescue. Could delete "Originating from variation within a host population" here and just add more information about what make evolutionary rescue special - that is the increase in locally adapted phenotypes. Bree comment. Minor – there are a lot of clauses in this sentence. Wonder if setting this as " – as necessary for evolution rescue – " might make the sentence more readable-->.
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals <!--# Bree comment: Again minor, but seems like this could be slimmed (as the sentence repeats the idea of promoting recovery twice and the "provide sources of individuals" seems awkward). It could read something like "In addition to promoting recovery of remnant populations after disease-induced declines, natural selection can also lead to increased resistance and/or tolerance…." -->with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->

There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
First, populations that survived <!--# Bree comment. Worth keeping tense consistent so that the four conditions feel parallel? This sentence could read "First, populations that survive following disease arrival evolve increased resistance" -->following disease arrival have evolved increased resistance.
Second, increased resistance contributes to higher individual fitness in the presence of disease <!--# "disease" or "the pathogen"? Resistance could convey complete prevention of disease, so I often struggle with which word to use -->and allows positive population growth.
Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
And fourth, reintroduced populations show long-term growth and persistence.
To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] <!--# Bree comment: Very minor but there are lots of places with long structurally complex sentences that seem just as good as broken up shorter sentences -->, and the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

@@

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"? Mark here:  This transition reads fine to me. -->.
We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

<!--# Bree comment: All of this was written before we knew the genomics had something cool to say. But now that the genomics feel a bit more elevated, I think this paragraph should be more bold. For example, we might want more here or in the Discussion to make it clear how the results of each of these sections are super important on their own. And that combining them is really amazing. I think the current tone is appropriately factual, but at some point we need to be louder about the novelty involved and the incredible amount of work that this represents.  This version feels like, we did a,b,c and so MYLF meet the conditions for feasibility (which makes it sound like we've only just gotten to the *starting* point of the study. I think we need to go further than saying they meet the four conditions of a specific conservation action. This is where beefing up the conclusions about evolutionary rescue seem appropriate. Also, genomic analyses typically come last in our papers as a dive into possible mechanistic explanations for observed patterns in the field. I think the order of the primary sections makes sense given the way you set up the argument about evolutionary rescue. That said, the order combined with the tone feels like it's contributing to the sense that the genomics are a box to be checked rather than contributing something exciting to the overall story. Now that Allie has added new analyses, I think we could make a slightly bolder set-up about how the parts fit together to make a surprisingly strong whole -->

<!--# Original pre-edit text from the above paragraph: Here, we report the results of (i) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd, (ii) a greatly expanded MYL frog reintroduction effort, and (iii) population viability modeling to assess the long-term persistence of reintroduced populations. Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian. These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases. -->

@@

To determine whether frog populations show an evolutionary response to Bd exposure, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$) <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi-diff.  -->.
In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overlapped for *F~ST~* and $\pi$ <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi - diff -->.
Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.

<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here. -->

Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
Our first analysis to detect signatures of selection identified a liberal set of variants that included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI; SI Dataset 1).
A single gene in both groups did not have an associated annotation.
For the SNPs in the stringent set of outliers, frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) <!--# Move this caveat text to the Discussion? :  and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors. -->, but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
No GO terms were significantly over-represented in the set of 29 annotated outlier genes.

Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? --> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9--> <!--# Allie comment: One is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows).  -->) and no GO terms were overrepresented within this set of genes.
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) <!--# Allie comment: For me this is enough of a reason to discuss this beyond the obvious relevance of the gene function.  -->, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.  -->and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove. Among susceptible species, survivors of epizootics bear alleles encoding these conformations."? What about MHC Class II characteristics linked to resistance by Bataille et al. (2015)? Savage and Zamudio (2011, 2016), Kosch et al. (2018)?-->

Together, our genomics results show that recovering populations have strong signatures of selection on multiple genes and regions of the genome.
While the functions of all genes and regions showing signatures of selection were not clearly interpretable, regions of the genome associated with resistant phenotypes in other amphibian systems such as the MHC complex, showed strong evidence of selection in recovering MYL frog populations.

@@

#### Frog translocations

@@

library(readr)
library(dplyr)
library(magrittr)
library(stringr)
library(lubridate)
d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))

@@

robust_year_max <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr" & secondary_period > 1) %>% 
  group_by(site_id) %>% 
  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
  summarize(max(max_2period))
```

Comparison of genomes of frogs from Bd-naive versus recovering populations (described above) indicated substantial differences in allele frequencies and genomic regions, consistent with strong selection following Bd exposure.
In this section, we describe the outcome of frog reintroductions in which animals for reintroduction were collected from recovering populations, including two of the five recovering populations in the frog evolution study (@fig-allelefrequencies A: population 1 and 4).
Reintroductions of *R. sierrae* were conducted as part of an effort to reestablish populations of this endangered species.
These reintroductions were conducted in Yosemite National Park (California, USA) and involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites [@joseph2018].
We refer to these reintroductions as "translocations" [@seddon2014].
During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap SI).
Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`) collected from 1 of `r n_distinct(d1$donor)` donor populations (@fig-yosemap SI).
Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].

<!--# To allow comparison with other translocation studies, add information about the proportion of translocated populations at which evidence of successful reproduction was observed in the 1-2 years following translocation. -->

To identify site and frog-specific drivers of post-translocation survival, we developed linear and multilevel models in a Bayesian framework.
Our specific interest was whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens SI).
In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens SI).
Collectively, these results show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance to minimize the adverse effects of Bd infection.
<!--# Allie comment: Maybe tolerance is a better word here? Since they were infected. Or maybe we need to explicitly define what we mean with resistance vs tolerance. Bree comment: I noticed this also – in the Intro, you do say that you are lumping both resistance and tolerance into the phrase "resistance" but you are talking about multiple things (being able to avoid infection, being able to fight infection, being able to tolerate infection, etc). Maybe at some point we want to address specifically that the genomic changes we are observing could contribute to multiple facets of resistance / tolerance.  And that we are lumping those together in an "evolved resistance" bin. Roland comment: We reworded this sentence, but probably have not fully dealt with resistance/tolerance issue. --> In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
Finally, in three locations where longer CMR time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
Therefore, frog resistance in non-natal habitats is maintained across generations.

Results from frog translocations indicated that adult survival was often relatively high and that nearly all populations showed evidence of successful reproduction (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).

>>>>

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth and prevents extinction [@carlson2014].
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022].
However, linking these selective responses directly to population recoveries - as necessary for evolutionary rescue - is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, the resistant/tolerant individuals <!--# How best to refer to resistance/tolerance? --> in recovering populations can also be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.

There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
First, populations that survive following disease emergence evolve increased resistance/tolerance.
Second, increased resistance/tolerance contributes to higher individual fitness in the presence of disease and allows positive population growth.
Third, the fitness benefits gained through selection for increased resistance are maintained beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
And fourth, reintroduced populations show long-term growth and persistence.
To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

@@

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in recovering populations [@knapp2016].
We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

To determine whether frog populations show an evolutionary response to Bd, we compared genomes of frogs in populations with different histories of Bd exposure.
Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@knapp2011, @vredenburg2010].

We used two complementary approaches to identify regions of the genome under selection.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific single nucleotide polymorphisms (SNPs).
Second, we used a splined window approach to identify larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).

The SNP analysis identified 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
One of the 7 identified genes did not have an associated annotation.
Frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.

The splined window analysis identified 58 outlier regions for *F~ST~* and 33 outlier regions for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2).
The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.  -->and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.

In summary, our genomics results show that recovering populations have signatures of strong selection on multiple regions of the genome.
Although the functions of all associated genes were not clearly interpretable, regions of the genome under selection in recovering populations included those associated with resistance/tolerance in other amphibian systems, such as the MHC complex.

@@

#### Frog population recovery

@@

library(readr)
library(dplyr)
library(magrittr)
library(stringr)
library(lubridate)
d1 <- read_csv(here::here("data", "clean", "frog_translocation_final.csv"))
d2 <- read_csv(here::here("data", "clean", "frog_counts_table.csv"))
d3 <- read_csv(here::here("data", "clean", "frog_captures_table.csv"))

@@

robust_year_max <- surveys %>% 
  mutate(yr = year(ymd(survey_date))) %>%
  filter(survey_type == "cmr" & secondary_period > 1) %>% 
  group_by(site_id) %>% 
  summarize(max_2period = max(secondary_period)) %>% ungroup() %>% 
  summarize(max(max_2period))

frog_counts_gt0_nrow <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  nrow()

frog_counts_gt0_median <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  summarize(median(pyears_tadsub_present)) %>% 
  round(2)

frog_counts_gt0_min <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  summarize(min(pyears_tadsub_present)) %>% 
  round(2)

frog_counts_gt0_max <- d2 %>% 
  filter(pyears_tadsub_present > 0) %>% 
  summarize(max(pyears_tadsub_present))

frog_captures_gt0_nrow <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  nrow()

frog_captures_gt0_median <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  summarize(median(pyears_recruits_present)) %>% 
  round(2)

frog_captures_gt0_min <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  summarize(min(pyears_recruits_present)) %>% 
  round(2)

frog_captures_gt0_max <- d3 %>% 
  filter(pyears_recruits_present > 0) %>% 
  summarize(max(pyears_recruits_present))
```

In this section, we describe the outcome of *R. sierrae* reintroductions conducted in Yosemite National Park as part of a 15-year effort to reestablish populations of this endangered species (`r min(d1$year)`--`r max(d1$year)`).
Each of the `r n_distinct(d1$site_id, d1$date)` reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) [@knapp2016] and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@joseph2018; @seddon2014].
The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).

Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].

Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or subadults.
For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
In summary, survey results suggest that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.

To determine whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection, we developed linear and multilevel models.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens SI).
Collectively, these results are consistent with frogs in recovering populations having evolved sufficient resistance/tolerance to minimize the adverse effects of Bd infection.
In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that frogs trasnslocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats.
Finally, in 3 locations where longer CMR time series allowed us to assess the survival of new adults recruited to the population, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
Therefore, frog resistance in non-natal habitats is maintained across generations.

#### Long-term population viability

Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).
